Topological Equivariant Artist Model for Visualization Library Architecture

### 2024-05-28 paper/main.tex
clean up structure, bring types in

\IEEEPARstart{V}{isualization} tools implement transforms of the form $data \rightarrow graphic$ and are expected to be structure preserving. Loosely, that means the mathematical properties of the data and how the points are connected to each other should be inferable such that  $graphic \rightarrow data$. For example, values read off a bar chart have to be equivalent to the values used to construct that chart. Therefore a visualization tool is structure preserving when it preserves the mapping $data\leftrightarrow graphic$.

@@

\begin{figure}[H]
  \includegraphics[width=1\columnwidth]{k_different_types.png}
  \caption{
  \note{add first panel which is table + all K type, second is scatter, etc...}
%
  \label{fig:related-work:continuity:ktypes}}
\end{figure}

@@

The structure on values has traditionally been codified using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical structure\cite{stevensTheoryScalesMeasurement1946} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale1971, thomasMathematizationNotMeasurement2014}.

 \textit{equivariance} allows us to test stucture preservation under transformations to data or visualization has been proposed by Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}.

 \textit{homomorphism} provides a way of asserting whether the relations between data elements was preserved as proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}.

\subsubsection{equivariance}
Specifically, Steven's conceptualizes the structure on values as \textcolor{action}{actions} on groups \footnote{A \textit{group} is a set with an associative binary operator. This operation must have an identity element and be closed, associative,
and invertable}:

@@

where the measurements are the groups and changes to them are actions. A function that preserves structure when the input or output is changed by a group action is called \textit{equivariant}. Given a group $G$ that acts on both the input $X$ and the output $Y$ of a function $f: X \rightarrow Y$
\begin{definition}\label{def:equivariance}
 A function $f$ is \textbf{equivariant} when $f(act(g,x)) = act(g,f(x))$ for all $g$ in $G$ and for all $x$ in $X$ \cite{pittsNominalSetsNames2013}
\end{definition}
which means that a visualization is structure preserving when there exist compatible group actions on the data and visualization, as proposed by Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}.

\subsection{homomorphism}

@@

\subsection{Topology}
\note{lead with homomorphism}
Borrowing structures from algebraic topology allows us to describe structure on individual elements in a way that holds true for the entire dataset, meaning we can use the same tools to check if structre is preserved regardless of whether the data fits in memory, is distributed, or is streaming. Our framework generalizesis because, informally, a topology $\mathcal{T}$ is a partitioning of a space into its subspaces in a way where those subspaces can be composed back into the space; the choice of partitions is how we encode the connectivity of the data and visualizations that we are interested in.

>>>>

\IEEEPARstart{V}{isualization} tools implement transforms of the form $data \rightarrow graphic$ and are expected to be structure preserving. Loosely, that means the mathematical properties of the data and how the points are connected to each other should be inferable such that  $graphic \rightarrow data$. For example, values read off a bar chart have to be equivalent to the values used to construct that chart. Therefore a visualization tool is structure preserving when it preserves the bidirectional mapping $data\leftrightarrow graphic$.

@@

@@

Data is often described by its mathematical structure, for example the Steven's measurement scales define nominal, ordinal, interval, and ratio data by the allowed operations on each \cite{stevensTheoryScalesMeasurement1946} and other researchers have since expanded the scales to encapsulate more types of structure \cite{leaFormalizationMeasurementScale1971, thomasMathematizationNotMeasurement2014}.

Loosely, the scales classify data as a set of values and the allowed transformations on that set, which can be operations, relations, or generalized as actions:

@@

Computer science data types roughly describe structure in an equivalent manner;
objects are elements of the set and class operators and methods are akin to
actions, as described in \autoref{sec:atct:fb:fiber}. Given this equivalancy between code and math, we can develop software functions to test whether structure is preserved based on mathematical tests of structure preservation. For example, \textit{equivariance} allows us to test stucture preservation under transformations to data or visualization and has been proposed by Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014} and \textit{homomorphism} provides a way of asserting whether the relations between data elements was preserved as proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}.

\subsubsection{equivariance}
Specifically, Steven's conceptualizes the structure on values as \textcolor{action}{actions} on groups \footnote{A \textit{group} is a set with an associative binary operator. This operation must have an identity element and be closed, associative,
and invertable}. A function that preserves structure when the input or output is changed by a group action is called \textit{equivariant}. Given a group $G$ that acts on both the input $X$ and the output $Y$ of a function $f: X \rightarrow Y$

\begin{definition}\label{def:equivariance}
 A function $f$ is \textbf{equivariant} when $f(act(g,x)) = act(g,f(x))$ for all $g$ in $G$ and for all $x$ in $X$ \cite{pittsNominalSetsNames2013}
\end{definition}
which means that a visualization is structure preserving when there exist compatible group actions on the data and visualization, as proposed by Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}.

\subsubsection{homomorphism}

@@

\subsection{Topology}

\begin{figure}[H]
  \includegraphics[width=1\columnwidth]{k_different_types.png}
  \caption{
  \note{add first panel which is table + all K type, second is scatter, etc...}
%
  \label{fig:related-work:continuity:ktypes}}
\end{figure}
\note{lead with homomorphism}
Borrowing structures from algebraic topology allows us to describe structure on individual elements in a way that holds true for the entire dataset, meaning we can use the same tools to check if structre is preserved regardless of whether the data fits in memory, is distributed, or is streaming. Our framework generalizesis because, informally, a topology $\mathcal{T}$ is a partitioning of a space into its subspaces in a way where those subspaces can be composed back into the space; the choice of partitions is how we encode the connectivity of the data and visualizations that we are interested in.
